---
title: "STATS 330 - ASSIGNMENT 1"
author: "jras611"
date: "09/03/2021"
output: word_document
---
# STATS330 - Assignment 1

## Question 1a

We will first examine the relationship between success and treatment using tables of counts and odds. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Reading in CSV file
stones.df = read.csv('stones.csv', header = TRUE)

```

### i) Creata a table of counts of success and treatmnt

```{r Question1ai Contingency Table}

#Question 1a.i
#Creating Contingency Table
stones.tbl = with(stones.df, table(success, treatment))
stones.tbl


```

### ii) Calculate the odds of successful removal of kidney stones for study participants receiving treatment A.

Probability is the fraction of times you expect to see a particular event in however many n total trials. So the probability of a succesful kidney stone removal for study participants receiving treatment A is the succesful kidney stone removals observed under treatment A divided by the total observations under treatment A.

The odds however are an expression of relative probabilities and defined as the probability that the event will occur divided by the probability that the event will not occur. If the probability of an event occuring is p and the probability of the same event not occurring is 1-p, then the odds of an event is p divided by 1-p (p/(1-p)).


```{r Treatment A Calculations}

#Probability of a succesful kidney stone removal under treatment A
success_A = 273/(273 + 77) 

#Odds of a succesful kidney stone removal under treatment A
success_A_odds = success_A / (1-success_A)
success_A_odds


```

Thus, there is a 78% probability of a succesful kidney stone removal for study participants receiving treatment A while the odds of a succesful kidney stone removal for study participants receiving treatment A is 3.5 successes to 1 failure.

### iii)Calculate the odds of successful removal of kidney stones for study participants receiving treatment B.

The same formula for probability and odds would apply for study participants under treatment B. 


```{r Question 1aii. Treatment B calculations}

#Probability of a succesful kidney stone removal under treatment B
success_B = 289/(61 + 289) 

#Odds of a succesful kidney stone removal under treatment B
success_B_odds = success_B / (1-success_B)
success_B_odds
```

Therefore, for study participants receiving treatment B the odds of success is 4.7 successes to 1 failures.

### iv)Calculate the ratio of the odds calculated in parts (ii) and (iii) and interpret inwords.  Which treatment is more successful for the removal of kidney stones?

The odds ratio of a 2-by-2 contingency table is the ratio of row-1 odds over row-2 odds. In plain words, the estimate of the odds ratio is the product of the diagonal values in the contingency table, divided by the product of the off-diaonal values


```{r Question 1aiv. Odds Ratio Calculations}
stones.tbl = with(stones.df, table(success, treatment))
stones.tbl

#Calculating Odds Ratio
odds_ratio = (77 * 289)/(273 * 61)
odds_ratio


```
The odds of a successful kidney stone removal  through treatment B is about 1.3 times higher than the odds for a successful kidney stone removal under treatment A. Therefore, the more succesful treatment for the removal of kidney stones is treatment B.


## Question 1b

### i) identify the response variable and the explanatory variable.

The response variable is success and the explanatory variable is treatment.

### ii) Justify and fit an appropriate model for this exploration.

I've chosen a Logistic regression model for this exploration as we are handling data with a Binomial distribution since we are assessing the probability of a Successful or Unsuccessful outcome in the operation that is repeated multiple times.

Furthermore, the Binomial distribution better reflects the nature of a discrete distribution, acknowledges that the number of successes associated with the ith observation must be an integer between 0 and $n_i$ and accounts for nonconstant variance.

$logit(p_i) = \beta_0 + \beta_1 x_i$

$Y_i$ ~ $Binomial(n_i, p_i)$

```{r Question 1b ii. Model fit}

#Fitting Logistic Regression Model
stones.fit = glm(success~treatment, data = stones.df, family = 'binomial')
  
summary(stones.fit)

```

```{r Coefficients of stone.fit}
exp(coef(stones.fit))

```

### iii) Interpret  the  model, being  sure to communicate  the uncertainty  in your interpretation

We can make the following statements:
- We estimate that the odds of succesful kidney stone removal for a study participant under treatment A is 3.5
- We estimate that for every study participant, the odds of a succesful kidney removal under treatment B is multiplied by 1.3

```{r Question 1biii) Interpretation}
exp(confint(stones.fit))

100*(exp(confint(stones.fit))-1)

```
-We estimate that for every study participant, the odds of succesful kidney removal under treatment B is multiplied by between 0.92 and 1.9 

- We estimate that for every study participant, the odds of a succesful kidney stone removal under treatment B is multiplied by between -8% and 94%.

### iv) How do the model results compare to your answer in (a), part(iv) above?

The model  results align with the answer to 1a(iv). As we found in question 1a(iv) the odds of a succesful kidney stone removal under treatment A is 3.5 and likewise, these odds are multiplied by 1.3 under treatment B.

## Question 1C

### i) Create two tables of counts of success and treatment, one for small kidney stones and one for large kidney stones.
```{r Question1ci Size Contingency Tables}
#Creating Tables for Large and Small succesful kidney stone removals
large.df = subset(stones.df, size == 'large')
small.df = subset(stones.df, size == 'small')

large.tbl = with(large.df, table(success, treatment))
small.tbl = with(small.df, table(success, treatment))

#Large Kidney Stones Table
large.tbl
#Small Kidney Stones Table
small.tbl


```
### ii) Calculate the odds of successful removal of small kidney stones for study participants receiving treatment A.

```{r Question 1cii) Small Odds Calculations A}
#Probability of Success for Small Kidney Stones under Treatment A
small_pA = 81/87
small_pA

#Odds of Success for Small Kidney Stones under Treatment A
small_oddsA = small_pA/(1-small_pA)
small_oddsA

```

The probability of successful removal of small kidney stones for study participants receiving treatment A is 0.93. 

As such, the odds of successful removal of small kidney stones for study participants receiving treatment A is 13.28571

### iii) Calculate the odds of successful removal of small kidney stones for study participants receiving treatment B.


```{r Question 1ciii) Small Odds Calculations B}
#Probability of Success for Small Kidney Stones under Treatment B
small_pB = 234/270
small_pB

#Odds of Success for Small Kidney Stones under Treatment B
small_oddsB = small_pB/(1-small_pB)
small_oddsB

```

The odds of successful removal of small kidney stones for study participants receiving treatment B is 6.5.

### iv) Calculate the ratio of the odds calculated in parts (ii) and (iii) and interpret in words. Which treatment is more successful for the removal of small kidney stones?

```{r Question 1civ) Small Odds Ratio}
#Small Kidney Stone Table
small.tbl
small_ratio = (6 * 234)/(81 * 36)
small_ratio

```


The ratio of odds for the succesful removal of small kidney stones under treatment B is 0.48 higher than the odds for succesful removal of small kidney stones under treatment A. The more succesful treatment for the removal of small kidney stones is treatment A.

### v) Calculate the odds of successful removal of large kidney stones for study participants receiving treatment A.

```{r Question 1cv) Large Odds Calculations A}
#Calculating the probability of a succesful removal of a large kidney stone under treatment A
large_pA = 192/263

#Calculating odds
large_oddsA = large_pA/(1-large_pA)
large_oddsA

```

The odds of a successful removal of large kidney stones for study participants receiving treatment A is 2.704225.

### vi) Calculate the odds of successful removal of large kidney stones for study participants receiving treatment B.

```{r Question 1cvi) Large Odds Calculations B}
#Calculating the probability of a succesful removal of a large kidney stone under treatment B
large_pB = 55/80

#Calculating odds
large_oddsB = large_pB/(1-large_pB)
large_oddsB

```

The odds of successful removal of large kidney stones for study participants receiving treatment B is 2.2

### vii) Calculate the ratio of the odds calculated in parts (v) and (vi) and interpret in words. Which treatment is more successful for the removal of large kidney stones?

```{r Question 1cv) Large Odds Ratio}
#Large Kidney Stone Removal
large.tbl
large_ratio = (71 * 55)/(192 * 25)
large_ratio

```
The ratio of odds for the succesful removal of large kidney stones under treatment B is 0.81 higher than the odds for succesful removal of large kidney stones under treatment A. The more succesful treatment for the removal of large kidney stones is treatment A.

## Question 1d

### Comment on what you discovered in (a) part (iv) and (c) parts (iv) and (vii). Are you surprised? If so, why? If not, why not?

In (a) part(iv) we discovered that the odds of a successful kidney stone removal  through treatment B is about 1.3 times higher than the odds for a successful kidney removal under treatment A.

We extrapolate on that in (c) parts (v) and (vi) and learn that the ratio of odds for the succesful removal of small kidney stones under treatment B is 0.48 higher than the odds for succesful removal of small kidney stones under treatment A and also that the ratio of odds for the succesful removal of large kidney stones under treatment B is 0.81 higher than the odds for succesful removal of large kidney stones under treatment A, and that in those specified cases of succesfully removing a small or large kidney stone; treatment A was actually the preferrable option. 

This is surprising to see considering that we discovered early on in Q1a (iv) and was supported through the code Q1b(iv) that the odds of success under treatment B was 1.3 times higher than treatment A. 

I believe this discrepancy could be due to the fact that treatment B is most typically used for small kidney stone removals with 270 out of 350 the study participants observed under treatment B (77%) undergoing a small kidney stone removal, as opposed to the majority of treatment A study participants (263 out of 350 (75%)) who had to have a large kidney stone removal. 

I imagine there is a significant difference in not only the procedure, but also the risk of failure or success among the two sizes of kidney stones during removal such that even if treatment A would be the more reliable option to remove either; small kidney stones are more likely to be easily and thus succesfully removed such that they are typically handled with treatment B since there is less risk for complication with smaller kidney stones. In that case, it makes sense why despite the findings of Q1c (v) and (vi), the data for treatment B may make it seem like the best treatment possible to ensure a succesful removal of a kidney stone.

## Question 2a

### Create appropriate plots to explore the following relationships, and comment on what the plots tell you:
```{r Question2a}
#Loading Species dataframe
species.df = read.csv('species.csv', header = TRUE)
names(species.df)[names(species.df) == 'ï..pH'] = 'pH'

#Subsetting each variable
Biomass = species.df$Biomass
pH = species.df$pH
Species = species.df$Species

#Biomass and pH
boxplot(Biomass ~ pH)
```
As Biomass is a numeric variable and pH is a categorical variable, it is better to explore the relationship of the two variables (Biomass and pH) through a boxplot. 

We can notice that the first boxplot for the group with a high pH level is comparatively tall, suggesting some significant difference across the other pH levels (low and mid) as well as significant variance in Biomass across the observations of the same high pH level. 

Whereas the boxplot of those with low pH levels is comparatively short suggesting less variance across the Biomass values across observations of the same low pH level.

Both the boxplots for high and low pH levels are relatively symmetrical.The mid pH level group appears to have a larger upper quartile than a lower quartile, indicating more variance in observations with greater Biomass values.

The minimum of each pH level group is approximately similar, whereas the maximum is a lot more varied. In fact, the median of the high pH level group is approximately the same as the maximum of the low pH group and the upper quartile of the mid pH group. 


```{r Biomass and Species}

#Biomass and Species
plot(Biomass, Species)

```
Biomass and Species are both numeric variables and thus can be plotted via a scatterplot. The following plot generally shows a downward trend between Species and Biomass thus indicating a negative relationship such that as Biomass increases, Species decreases.

```{r Species and pH}

#Species and pH
boxplot(Species ~ pH)


```
Species is a numeric variable while pH is a categorical variable, therefore a boxplot would be most appropriate to explore their relationship. Immediately, we notice a resemblance to the Biomass and pH plot,.The group with high pH tends to have a greater number in species than those of low or mid pH levels. It appears that there is a great difference in maximums between species count of high pH level observations to low and mid pH observations, that the relationship is relatively symmetrical as well and that the median of the high pH level boxplot is the same as the upper quartile of the mid pH level boxplot.

The similar minimums also bears resemblance, except for the low pH level plot which has a much lower minimum and a shorter plot with perhaps a larger lower quartile than upper quartile indicating more variance in the observations of low pH level with small species count.

Differences also lie in how the mid pH level box plot is more symmetrical for species.


## Question 2b

### Create a plot that explores the relationship between Biomass, Species and pH. Comment on what the plot tells you.

```{r Question 2b}
#Colouring plot by pH level
group.cols <- c(high = "black", mid = "blue", low = "red")
point.cols <- group.cols[as.character(pH)]
#Plotting
plot(Biomass, Species, col = point.cols)
#Adding Legend
legend("topright", legend = c("High", "Mid", "Low"), pch = rep(1, 3), col = c("black", "blue", "red"), title = 'pH level')


```
From the plot we can notice that Biomass and Species has a generally negative relationship where as Biomass increases, Species decreases, but in addition to the classification by pH level we can derive the insight that observations with High pH levels tend to also have a greater amount of Species and a larger value for Biomass and that the amount of Species or Biomass tends to decrease as pH level decreases. 

## Question 2c

### i) Identify the response and explanatory variables for this exploration.

As we are interested in exploring the relationship between the condition of the soil and the number of plant species, the response variable would be the number of plant species and the explanatory variables would be the condition of the soil measured by Biomass * pH.

### ii) What model would be appropriate to fit? Choose from Linear regression, Poisson regression and Logistic regression. Justify your choice

As the response variable consists of discrete count data, the appropriate model to fit is a Poisson regression. A Poisson regression model assumes a discrete distribution for the given discrete response, accounts for nonconstant variance which we suspected in our investigation of the plots and relationships in 2a and 2b and ensures that the epected value is greater than 0 for all observations. Thus, Poisson regression would be the best choice for this model fit.

### iii) Using your answers to parts (i) and (ii), run the following code for two models making sure you replace the blanks with the appropriate terms. Briefly interpret the 
$models^3$.

Model 1
```{r Question 2ciii Model1}
model1 = glm(Species~Biomass*pH, family = "poisson", data = species.df)

summary(model1)
```
In other words, we have
$$ \mu_i = \exp(\beta_0 + \beta_1b_i + \beta_2m_i + \beta_3l_i + \beta_4b_im_i + \beta_5b_il_i), $$
where $b_i$ is the biomass of the $i$th soil, and $l_i$ and $m_i$ are dummy variables for the low- and mid-pH levels, respectively.

There are two kinds of effects: 
The main effects of Biomass and the respective pH levels which only involve a single variable and the interaction effects such as the interaction between Biomass and pH level which involves more than one variable.

We can interpret the main effects in the same way as we can for models that do not have an interaction as the baseline level is another variable which is a factor.

```{r Model 1 Interpretation}
#Interpreting Main effect of Biomass
100*(exp(10*confint(model1)[2, ])-1)

```

We estimate that, for every 10m/a increase in the biomass of soil, the expected number of plants in species is associated with a decrease of between 73.21439% and 56.28961%.

```{r Model 1 Interpretation 2}
#Interpreting Low pH levels
100*(exp(confint(model1)[3, ])-1)
```

We estimate that low levels of pH is associated with a decrease in the expected number of plant species of between 63.89244% and 45.95685%.

```{r Model 1 Interpretation 3}
#Interpreting Mid pH levels
100*(exp(confint(model1)[4, ])-1)

```

We estimate that mid levels of pH is associated with a decrease in the expected number of plant species of between 40.09656% and 14.01968%.

```{r Model 1 Interpretation 4}

100*(exp(confint(model1)[5, ])-1)
100*(exp(confint(model1)[6, ])-1)

```

An interaction effect, measures how the effect of one variable changes depending on the other. $\beta_4$ and $\beta_5$ measures the differences in the intercept and slope respectively between the mid level pH and high level pH and the low level pH and high level pH respectively. From the calculation above we can estimate that the effect of mid pH level decreases by between approximately 21% and 7%, whilst the effect of low pH level changes from between a 7% decrease to a 1% increase. 

Model 2
```{r Model 2}
model2 = glm(Species~Biomass+pH, family = "poisson", data = species.df)
summary(model2)

```

In other words, we have
$$ \mu_i = \exp(\beta_0 + \beta_1*w_i + \beta_2*h_i + \beta_3*l_i), $$
where $b_i$ is the biomass of the $i$th soil, and $l_i$ and $m_i$ are dummy variables for the low- and mid-pH levels, respectively.

As opposed to Model 1. Model 2 fits an interaction effect which measures how the effect of one variable changes depending on the other. In this case, the high pH level group.

```{r Model 2 Intepretations}

100*(exp(confint(model2))-1)


```
Holding the pH of the soil constant, we estimate that the number of plant species in a plant plot 
- in the mid pH level group is between 42$ and 28% lower than an observation in the high pH level group.
- in the low pH level group is between 71% and 63% lower than an observation in the high pH level group

### iv) Evaluate the adequacy of the models fitted in part (iii). Comment briefly

From the plot, we can discern that the residual deviance of model 1 is 83.201 whereas the degrees of freedom is 84. Having the residual deviance be smaller than the degress of freedom is typically a good sign for adequacy.

In comparison, model 2 has a residual deviance of 99.242 on 86 degrees of freedom so we can infer that model 1 would likely be the more adequate model to fit.

### v) Calculate the change in residual deviance from model1 to model2 and calculate the  probability of observing such a change given we have ‘lost’ 2 degrees of freedom (i.e., we have reduced the number of parameters in the model from 6 to 4).

We know from the summary() function performed for each model in Q2 iii) that the residual deviance for model 1 and model 2 is 83.20114 and 99.2415 respectively.
```{r Model Deviances}
#model1 deviance
print('Model 1 Deviance')
dev1 = deviance(model1)
dev1
#model2 deviance
print('Model 2 Deviance')
dev2 = deviance(model2)
dev2
```

Since we know that model 1 deviance < model 2 deviance, we proceed to calculate the change in residual deviance from model 2 to model 1.

```{r Residual Deviance Change}
#residual deviance calculation
dev2 - dev1

#or alternatively
anova(model1, model2)

```
The change in residual deviance is 16.04036.


### vi) Construct a plot displaying the estimated effects of your preferred model (i.e. model1 or model2) on the response $variable^4$. Clearly communicate what your plot displays.
```{r Estimation Plot}

plot(Biomass, Species, col = point.cols, main = 'Estimated Effects of Biomass and pH on Species')
## A vector of 90 Biomass values between 0.05 m/a and 10m/a.
xx <- seq(0.05, 10, length.out = 90)
## A data frame for the high pH level group.
new.data.high <- data.frame(Biomass = xx, pH = rep("high", 90))
## A data frame for the low pH level group.
new.data.low <- data.frame(Biomass = xx, pH = rep("low", 90))
## A data frame for the mid pH level group.
new.data.mid<- data.frame(Biomass = xx, pH = rep("mid", 90))

## Fitted values for each of the three groups.
yy.mid <- predict(model1, new.data.mid, type = "response")
yy.low <- predict(model1, new.data.low, type = "response")
yy.high <- predict(model1, new.data.high, type = "response")
## Adding lines to the plot.
lines(xx, yy.high, col = "black")
lines(xx, yy.low, col = "red")
lines(xx, yy.mid, col = "blue")
## Adding the legend.
legend("topright", legend = c("High", "Mid", "Low"), pch = rep(1, 3), col = c("black", "blue", "red"), lty = rep(1, 3))

```
The plot above shows the estimated effects of model 1 on the response variable Species. In particular, it estimates that as Biomass increases, Species will decrease and the separate lines across the pH levels distinguishes the estimated effect of Biomass on species by pH, with each respective pH generally following the same downward trend albeit at differring values of Biomass and Species corresponding to their pH level (High pH Level = Greater number of Biomass and Species vs. Low pH level = Smaller number of Biomass and species)

### vii) Based on your model, calculate and interpret 95% confidence intervals for the mean number of plant species observed in each of three separate plots, all of which have Biomass=5 but each with different levels of pH: low, mid, and high.

```{r}

#Creating new data with given conditions
newdata = data.frame(Biomass = 5, pH = c("high", "mid", "low"))
#Predicting
predict.glm(model1, newdata = newdata, type = "response", se.fit = TRUE)


```

The average number of plant species observed in three separate plots of 5m/a in Biomass and differing pH levels (high, mid and low) is estimated to be: *25 (0dp) with  a standard error of 0.9 for observations with high pH levels
*16 (0dp) with a standard error of 0.9 for observations with mid pH levels
* 5 with a standard error of 0.7 for observations with low pH levels.